---
title: "data_manipulation"
output: github_document
date: "2025-09-18"
---

```{r}
library(tidyverse)
library(readxl)
```

Lets import the datasets.

```{r}
litters_df <- 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", ""))
litters_df <- 
  janitor::clean_names(litters_df)
```

```{r}
pups_df <- 
  read_csv("data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3)
pups_df <- 
  janitor::clean_names(pups_df)
```

## 'select'

'select' selects variables from my dataframe.

Select specific variables.
```{r}
select(litters_df, group, litter_number, gd0_weight)
```

Select a range.
```{r}
select(litters_df, group:gd_of_birth)
```

Select by removal.
```{r}
select(litters_df, -group)
```

```{r}
select(litters_df, -(group:gd_of_birth))
```

Select by prefix.
```{r}
select(litters_df, group, starts_with("gd"))
```

Rearrange by selecting.
```{r}
select(litters_df, starts_with("gd"), group)
```

```{r}
select(litters_df, litter_number, everything())
```

it does the same thing:
```{r}
relocate(litters_df, litter_number)
```

Rename by selecting.
```{r}
select(litters_df, GROUP = group, everything())
```

or:
```{r}
rename(litters_df, GROUP = group)
```

What if I only select one variable?
```{r}
select(litters_df, group)
```

Compare with:
```{r}
pull(litters_df, group)
litters_df$group
```


## 'filter'

use 'filter' to filter out rows.

```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than 2 pups alive

```{r}
filter(litters_df, pups_born_alive > 2)
```

look for non-matches

```{r}
filter(litters_df, pups_born_alive != 6)
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with 'drop.na()'

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```


## 'mutate'

use 'mutate()' to modify or create variables

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```


## 'arrange'

'arrange()' arranges things

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, group, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))
```


## Pipes

Use pipes to avoid clunky stuffs

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) %>% 
  janitor::clean_names() %>% 
  select(group, litter_number, starts_with("gd")) %>% 
  drop_na() %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```














